Application of proper orthogonal decomposition to measurements of steady engine head flow

### 2016-09-24 pod_encit/trunk/paper.tex
new lib version

\documentclass[10pt,fleqn,a4paper]{article}
\usepackage{encit2010}
\usepackage{amsmath}
\usepackage{amsthm}

@@

% LEGEND FOR THE GEOMETRIES
% Action = Head 1
% Scania = Head 2
% Sygma = Head 3
% Ducato = Head 4

\begin{document}
\hspace{-8.5mm}
\begin{tabular}{||p{\textwidth}}
\begin{center}
\vspace{-4mm}
\title{ANALYSIS OF STEREOSCOPIC PIV MEASUREMENTS OF IN-CYLINDER FLOWS USING PROPER ORTHOGONAL DECOMPOSITION}
\end{center}
\authors{Rodrigo Piccinini, rodrigo.piccinini@sygmamotors.com.br} \\
\authors{Carla Fernandes, carla.fernandes@sygmamotors.com.br} \\
\authors{Marcelo Andreotti, marcelo.andreotti@sygmamotors.com.br} \\
\institution{Sygma Motors, Av. Cassiano Ricardo, 1306, sala 2, Jardim Alvorada, CEP 12240 - 540, S\~{a}o Jos\'{e} dos Campos - SP - Brasil} \\
\\
\abstract{\textbf{Abstract}

% SHORT ABSTRACT

%In the present work, the Proper Orthogonal Decomposition (POD) was applied in the
% identification of flow coherent structures. The experiment consisted of 
% mounting on an Plexiglas cylinder the intake ports of an internal 
% combustion engine and establishing a steady flow through this system.
%Velocity fields at different cylinder cross sections and for several 
%instants of time were measured by Particle Image Velocimetry (PIV).
%It was shown that the principal eigenvector might contains up to 80\% of the
%time averaged kinetic energy of the flow field.

In the present work, the Proper Orthogonal Decomposition (POD) was applied in the identification of flow coherent structures. For the velocity flow field, stereo Particle Image Velocimetry (PIV) measurements were carried out in a fixed cross section. We have observed that the principal eigenvector contains up to 80$\%$ of the time averaged kinetic energy of the flow field.
}
\\
\\
\keywords{\textbf{Keywords:} proper orthogonal decomposition, stereo PIV, coherent structures, turbulent flow, IC engines.}\\
\end{tabular}

\section{INTRODUCTION}

In the development of cylinder head intake ports for spark-ignited internal combustion engines, the characterization of coherent flow structures generated during induction - the most usual being tumble and swirl motions - is an important issue for combustion acceleration.

The flow field will directly affect the mixture formation and hence the spatial fuel distribution inside the combustion chamber. A spatially non-homogeneous mixture leads to changes in flame behavior in terms of heat release, propagation speed, etc. This behavior is not well understood and has motivated the research reported in numerous studies (Haworth et al., 2000; Jim\'{e}nez et al., 2002; Pasquier et al., 2007; H\'{e}lie and Trouv\'{e}, 2000; Xiong and Roberts, 2002). The characteristics of the combustion depend strongly on local variations of mixture and also on the length-scale of these variations.

In a general way, higher turbulence levels during flame development reduce combustion duration.  A common strategy to achieve these higher turbulence levels consists of disintegrating large flow structures into turbulence near the spark release event. In this way, the kernel formation will depend on the local equivalence ratio and temperature around the spark plug. The flow field might also advect the initial kernel and the local turbulence will influence its evolution to a turbulent flame. The turbulent flame for the premixed thin reaction sheet will grow faster than the laminar flame due to the large scale turbulence ($l_{t}/\delta^{0}_{L} > 1$, where $l_{t}$ is the integral length scale and $\delta^{0}_{L}$ is the laminar flame thickness (Poinsot and Veynante, 2005)). A faster combustion event is desired because it will increase the efficiency of converting chemical energy to piston work. Therefore, the majority of combustion strategies will attempt to create such desirable high turbulence levels near the ignition.

For this purpose, the creation of a strong swirling flow inside the combustion chamber is often studied. The idea is to store a great amount of kinetic energy in large structures during the intake. During the compression, these structures will remain until the ignition event and then they will break into smaller structures increasing the turbulence level. In case of swirling structures not achieved during the intake, the kinetic energy of the flow might dissipate faster.

Furthermore, the turbulent flow is characterized by random and chaotic three-dimensional motion. In this way, the flow will differently affect the formation of large structures cycle-by-cycle, rendering the engine operation difficult. Thus, not only the magnitude of the swirling motion is important, but also its regularity. Therefore, the knowledge of the bulk motion in the cylinder and the turbulence evolution is of great importance for the improvement of engines.

The objective of the present work is the study of the statistical properties of the flow induced by the intake ports in four different cylinder heads using proper orthogonal decomposition (POD). In order to obtain the velocity flow field, the stereo particle image velocimetry (PIV) measurements were carried out in a specific cylinder cross section. This technique enables spatially resolved measurements of the three instantaneous components of the flow velocity field within a very short time and allows the detection of large and small scale spatial structures. Once the velocity field was acquired, the POD was applied in order to know how the flow energy is stored. This statistical analysis demonstrates a promising tool in the design of intake ports.

@@

The characterization of coherent structures is an interesting issue in the study of turbulent flows and it is the subject of vast literature (Heywood, 1988; Cosadia et al., 2006; Cazemier, 1997; Bergmann and Cordier, 2003, Hung and Hien, 1998; Boree, Maurel and Bazile, 2002; Li Y. et al., 2004). However, the mathematical definition of such structures is not necessarily standard and might differ depending on context. One possible way is the Proper Orthogonal Decomposition (POD) (Lumley, 1967), which establishes a criterium based on maximum energy content.

@@

The observations can be the result of experimental measurements or computer simulations. These results may be not necessarily the flow velocity, but any of the flow properties. Furthermore, each observation must be taken in a different instant of time. In the present work, one observation is made up of the full procedure required by the PIV technique (as described in the Section 2) to provide a velocity field in some instant of time. The frames are taken at different times equally spaced and each event is assumed linearly independent.

@@

According to Eq.~\eqref{eq: max-cond}, $\bv{\Phi}$ can not be zero because his norm equals one. Therefore:

@@

After performing the integrations and having the matrix $C$ (Eq.~\eqref{cov1}), the computation of eigenvalues and eigenvectors is carried out with the opensource library SciPy.

@@

The Figure~\ref{fig: energy_fraction} shows the content of time average energy in modes. The energy fraction content in each mode is presented by Fig.~\ref{fig: energy_fraction}(a). The Figure~\ref{fig: energy_fraction}(b) shows the cumulative energy when the modes are grouped from $\bv{\Phi}_1$ to $\bv{\Phi}_{100}$. For higher valve lifts, the flow field has shown more organized in the sense that more energy is stored in few modes. Only the first mode, $\bv{\Phi}_1$, contains up to 80$\%$ of the total energy (see the zoom of the Fig.~\ref{fig: energy_fraction}(a)). Moreover, from the 30$^{th}$ mode, the energy decreases following an exponential decay. The exponential law is given by $E(k) \approx a_1 e^{a_2 k} ,  k \ge 30$. The exponential constant ($a_2$) is in the interval of $[-0.031,-0.023]$ for all cases.

@@

\begin{figure}[h]
 \centering
 \includegraphics[width=0.78\textwidth]{./imgs/scania-modes.png}
 % scania-modes.png: 2196x879 pixel, 72dpi, 77.46x31.01 cm, bb=
 \caption{Velocity fields of Head 2: $\bv{\Phi}_1$ (left), $\bv{\Phi}_{15}$ (center) and $\bv{\Phi}_{30}$ (right).}
 \label{fig: organization}
\end{figure}

The link between energy content and the flow organization is supported by the observation of the mode velocity field presented in Fig.~\ref{fig: organization}. Here, the vector magnitudes are normalized according to the same norm defined in Eq.~\eqref{eq: max-cond}. It is possible to notice that the motion organization decreases with the higher order modes. Furthermore, these modes presents several vortical structures while the first mode contains the majority of swirling structure. Hence, $\bv{\Phi}_1$ have the majority of the angular momentum.

So far, only the time averaged energy has been discussed. However, if the $b^{s}_{j}$ coefficients from Eq.~\eqref{eq: Us} are computed, the energy time history is recovered. The Figure~\ref{fig: energy_hist} presents the energy history for the first mode and the set of the first ten modes. It is seen that the first mode is in general more oscillatory when viewed separately in the gray line.  This is rather consistent since energy transfers between the modes are expected. The energy decrease in the first mode possibly increases the energy of some other mode in the set and the other way around. This energy transfer will reduce the oscillations.

The Figure~\ref{fig: Lz} shows the contribution of each mode to angular momentum around the cylinder axis. It is observed that the first mode contains the main swirling motion while the subsequent modes corresponds to the oscillating motion as shown in Fig.~\ref{fig: organization}. As one could expect, the low valve lift of 2\,mm did not clearly define swirl motion because of great magnitude oscillations.

In order to define which intake port geometry will better perform, in terms of increase the turbulence level close to spark ignition event, not only the energy in the modes most be analyzed but also their respective angular momentum. As showed in Fig.~\ref{fig: energy_hist}, the results (c) and (d) have a similar energy history level in the first mode and for the set of ten modes. It is mean that the Head 2 and 3 distribute the energy in a similar way. This same trend is also observed in terms of energy spectra (see Fig.~\ref{fig: energy_fraction}). However, the angular momentum is quite different as presented in Fig.~\ref{fig: Lz} by the plots (c) and (d). Comparing the results of the first mode velocity field presented by Fig.~\ref{fig: scaniavssygma}, it is noticed that the Head 3 velocity field is composed by two-vortex structure against only one big vortex in the Head 2. This is the main reason in the angular momentum production.

@@

In order to better understand how the intake ports geometry influence the flow which enters inside the combustion chamber during the intake phase, a statistical decomposition, like proper orthogonal decomposition, was applied on the three dimensional velocity flow field obtained with a stereoscopic PIV.

The results have shown that the first mode contains up to 80$\%$ of the time average kinetic energy of the flow field and have the major swirling structure. Therefore, the first mode defines the average angular momentum. The other modes, which we can call low energy modes, have shown several vortical structures, as presented in Fig.~\ref{fig: organization}, and have an average angular momentum around zero (see Fig.~\ref{fig: Lz}).

In what concerns the flow regularity, Fig.~\ref{fig: Lz} shows that the first mode basically defines the average angular momentum, which is superposed by the contributions of higher modes with average angular momentum around zero. Looking for the differences from one port to another, $\bv{\Phi}_1$ is much more disturbed by higher modes in the case of Head 3. This implies in having more dispersed swirling structures in the engine when concerning the ensemble of different cycles. It is also notable the regularity of Head 4, for which all oscillations are rather small.

Even if a mean swirl is measured, we have shown that the structure of the mean flow is complex and difficult to interpret. Thus, POD has shown itself to be an important methodology to analyze the flow and an important tool for the design of the cylinder head.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{setspace}

@@

\title{Application of proper orthogonal decomposition to measurements of steady engine head flow}
\author{Rodrigo Piccinini, Carla Fernandes, Marcelo Andreotti.}

\begin{document}
\maketitle

\doublespacing

\abstract{The proper orthogonal decomposition (POD) is applied to particle image velocimetry (PIV) datasets of an engine static flow bench. Different flow fields are produced changing intake port design and flow snapshots are taken at a fixed position downstream port exit. It is observed that a swirl motion is mostly contained in the first POD mode and that the swirl motion steadiness depends on the amount of kinetic energy in the first mode. All higher POD modes showed pairs of counter rotating vortices with zero mean angular momentum.
}

\section{Introduction}

Characterization and understanding of flow field inside internal combustion engines is of primary importance to improvements in engine efficiency and predictability. Flow field directly affects mixture formation and turbulence level, both important factors dictating flame front advance.

In the development of intake ports for spark-ignited internal combustion engines, the characterization of coherent flow structures generated during intake - most usual ones being tumble and swirl motions - is an important issue for combustion acceleration.

The objective of the present work is the study of the statistical properties of the flow induced by the intake ports in four different cylinder heads using proper orthogonal decomposition (POD). In order to obtain the velocity flow field, the stereo particle image velocimetry (PIV) was performed in a specific cylinder cross section. This technique enables spatially resolved measurements of the three instantaneous components of the flow velocity field within a very short time and allows the detection of large and small scale spatial structures. Once the velocity field was acquired, the POD was applied in order to know how the flow energy was stored. This statistical analysis suggests to be a promising tool in the design of intake ports.

@@

Characterization of coherent structures is an interesting issue in the study of turbulent flows and it is the subject of vast literature (Heywood, 1988; Cosadia et al., 2006; Cazemier, 1997; Bergmann and Cordier, 2003, Hung and Hien, 1998; Boree, Maurel and Bazile, 2002; Li Y. et al., 2004). However, the mathematical definition of such structures is not necessarily standard and might differ depending on context. One possible way is the Proper Orthogonal Decomposition (POD) (Lumley, 1967), which establishes a criterium based on maximum energy content.

@@

The observations can be the result of experimental measurements or computer simulations. These results may be not necessarily the flow velocity, but any of the flow properties. Furthermore, each observation must be taken in a different instant of time. In the present work, one observation is made up of the full procedure required by the PIV technique (as described in the Section 2) to provide the velocity field in some instant of time. The frames are taken at different times equally spaced and each event is assumed linearly independent.

@@

According to Eq.~\eqref{eq: max-cond}, $\bv{\Phi}$ can not be zero because its norm equals one. Therefore:

@@

After performing the integrations and having the matrix $C$ (Eq.~\eqref{cov1}), the computation of eigenvalues and eigenvectors is performed with the opensource library SciPy.

@@

The Figure~\ref{fig: energy_fraction} shows the content of time average energy in modes. The energy fraction content in each mode is presented by Fig.~\ref{fig: energy_fraction}(a). The Figure~\ref{fig: energy_fraction}(b) shows the cumulative energy when the modes are grouped from $\bv{\Phi}_1$ to $\bv{\Phi}_{100}$. For higher valve lifts, the flow field is more organized in the sense that more energy is stored in fewer modes. Only the first mode, $\bv{\Phi}_1$, contains up to 80$\%$ of the total energy (see the zoom of the Fig.~\ref{fig: energy_fraction}(a)). Moreover, from the 30$^{th}$ mode, the energy decreases following an exponential decay. The exponential law is given by $E(k) \approx a_1 e^{a_2 k} ,  k \ge 30$. The exponential constant ($a_2$) is in the interval of $[-0.031,-0.023]$ for all cases.

@@

\begin{figure}
 \centering
 \includegraphics[width=0.78\textwidth]{./imgs/scania-modes.png}
 % scania-modes.png: 2196x879 pixel, 72dpi, 77.46x31.01 cm, bb=
 \caption{Velocity fields of Head 2: $\bv{\Phi}_1$ (left), $\bv{\Phi}_{15}$ (center) and $\bv{\Phi}_{30}$ (right).}
 \label{fig: organization}
\end{figure}

The link between energy content and the flow organization is supported by the observation of the mode velocity field presented in Fig.~\ref{fig: organization}. Here, the vector magnitudes are normalized according to the same norm defined in Eq.~\eqref{eq: max-cond}. It is possible to notice that the motion organization decreases in the higher order modes. Furthermore, these modes presents several vortical structures while the first mode contains one major swirling structure. It will be shown later that $\bv{\Phi}_1$ contains the greatest amount of the angular momentum.

So far, only the time averaged energy has been discussed. However, if the $b^{s}_{j}$ coefficients from Eq.~\eqref{eq: Us} are computed, the energy time history is recovered. The Figure~\ref{fig: energy_hist} presents the energy history for the first mode and the set of the first ten modes ($\bv{\Phi}_1 + \bv{\Phi}_2 + ... +\bv{\Phi}_{10}$). It is seen that the first mode is in general more oscillatory when viewed separately in the gray line.  This is rather consistent since energy transfers between the modes are expected from the nonlinearity of the Navier-Stokes equations. The energy decrease in the first mode possibly increases the energy of some other mode in the set and the other way around. This energy transfer reduces the oscillations of the set.

The Figure~\ref{fig: Lz} shows the contribution of each mode to the angular momentum around the cylinder axis. It is observed that the first mode contains the main swirling motion and the subsequent modes correspond to the oscillations shown in Fig.~\ref{fig: organization}. As one could expect, the low valve lift of 2\,mm did not clearly define a swirling motion because of the great magnitude oscillations.

As showed in Fig.~\ref{fig: energy_hist}, the results (c) and (d) have a similar energy history level in the first mode and for the set of ten modes. It means that the Heads 2 and 3 distribute the energy in a similar way. This same trend is also observed in the  energy spectra (see Fig.~\ref{fig: energy_fraction}). However, the angular momentum is quite different as presented in Fig.~\ref{fig: Lz} by the plots (c) and (d). Comparing the results of the first mode velocity field presented by Fig.~\ref{fig: scaniavssygma}, we see that the Head 3 velocity field is composed by a two-vortex structure against only one big vortex in the Head 2. This would explain the different angular momenta.

@@

A mathematical tool called Proper Orthogonal Decomposition was used along with the experimental technique of stereoscopic PIV in the study of the in-cylinder flow generated by the intake ports of an internal combustion engine.

The results have shown that the first mode contains up to 80$\%$ of the time average kinetic energy of the flow field and have the major swirling structure. In what concerns the flow regularity, Fig.~\ref{fig: Lz} shows that the first mode basically defines the average angular momentum, which is superposed by the contributions of higher modes with average angular momentum around zero. Looking for the differences from one port to another, $\bv{\Phi}_1$ is much more disturbed by higher modes in the case of Head 3. This implies in having more dispersed swirling structures in the engine when concerning the ensemble of different cycles. It is also notable the regularity of Head 4, for which all oscillations are rather small.

Even if a mean swirl is measured, we have shown that the structure of the mean flow is complex and difficult to interpret. Thus, POD has shown itself to be an important methodology to analyze the flow and an important tool in the design of the cylinder head.
